---
title: "Aneuploidies report"
author: "Nidia Barco Armengol"
date: "`r Sys.Date()`"
output: html_document
params:
  pool_data: NULL
  control_data: NULL
  synthetic_data: NULL
  synthetic: NULL
  best_norms: NULL
  threshold_signif: NULL
  output_dir: NULL
---

```{r create_plots, echo=FALSE, include=FALSE}

library("dplyr")
library("ggplot2")
library("ggrepel")
library("ggpubr")
library("circlize")
library("ggfortify")
library("vegan")
library("karyoploteR")
library("GenomicRanges")
library("patchwork")
library("ggforce")
library("plotly")

Pool_Rdata <- params$pool_data
load(Pool_Rdata)

data_pool <- subset(pool_coverage_values, pool_coverage_values$autosomes_mean>=50)

Control_Rdata <- params$control_data
load(Control_Rdata)

data_control <- subset(pool_coverage_values, autosomes_mean >= 50)

Synthetic_data <- params$synthetic_data
load(Synthetic_data)

altered_aneu <- data.frame(matrix(nrow = 0, ncol = 2))
colnames(altered_aneu) <- c("Sample", "Altered_autosome")

Chromosomes <- c(paste("Chr", 1:22, sep = ""))
signif_synth <- read.csv2(params$synthetic)
percentages <- seq(10, 100, by = 10)
best_norms <- read.csv2(params$best_norms)
plotlis <- list()

## Autosomes

for(c in Chromosomes){
  
  synthetic <- data_synth[grepl(paste0("_", c,"_"), data_synth$Sample),]
  synthetic_signif <- subset(synthetic, Sample %in% signif_synth$Synthetic_significant)
  synthetic_exclude <- subset(synthetic, !(Sample %in% signif_synth$Synthetic_significant))
  data_complete <- rbind(data_control, synthetic, data_pool)
  
  crom_synth_sign <- subset(signif_synth, signif_synth$Chromosome==c)
  
  for (i in 1:length(percentages)) {
  num <- percentages[i]
  mono_pattern <- paste0("Monosomy", num, "$")
  tri_pattern <- paste0("Trisomy", num, "$")
  
    if (any(grepl(mono_pattern, crom_synth_sign$Synthetic_significant)) & any(grepl(tri_pattern, crom_synth_sign$Synthetic_significant))) {
      sample1 <- crom_synth_sign$Synthetic_significant[grep(mono_pattern, crom_synth_sign$Synthetic_significant)]
      sample2 <- crom_synth_sign$Synthetic_significant[grep(tri_pattern, crom_synth_sign$Synthetic_significant)]
      signif_line <- c((i - 1) * 2 + 1, (i - 1) * 2 + 2)
      signif_percentage <- paste0(num, "%")
      break
    }
  }
  
  norms <- subset(best_norms, best_norms$Chromosome==c)
  data_norm_list <- list()
  
  for(n in norms$Normalization){
    if (n=="Norm_all"){
      Norm <- data.frame("Samples" = data_complete$Sample, 
                         "Normalized_values" = data_complete[[paste0(c, "_mean")]]/data_complete[[paste0("No", gsub(".*Chr(\\d+)", "\\1", c), "_mean")]])
      data_norm_list[[length(data_norm_list)+1]] <- Norm
    } else{
      Norm <- data.frame("Samples" = data_complete$Sample, 
                         "Normalized_values" = data_complete[[paste0(c, "_mean")]]/data_complete[[paste0(strsplit(n,"_")[[1]][2], "_mean")]])
      data_norm_list[[length(data_norm_list)+1]] <- Norm
    }
  }
  
  crom <- data.frame(data_norm_list[[1]]$Normalized_values, 
                     data_norm_list[[2]]$Normalized_values, 
                     data_norm_list[[3]]$Normalized_values)
    
  rownames(crom) <- data_norm_list[[1]]$Samples
  colnames(crom) <- norms$Normalization
    
  pca.filt <- prcomp(crom, scale=T)
  pca.data <- pca.filt$x
  
  ## Plotly
  
  pca.data <- subset(pca.data, !(rownames(pca.data) %in% synthetic_exclude$Sample))
  
  synth_colors <- rep(c("blue", "red"), 10)[nrow(synthetic_exclude)+1:20]
  colores <- c(rep("black", nrow(data_control)), synth_colors, rep("purple", nrow(data_pool)))
  
  fig <- plot_ly(data = as.data.frame(pca.data), x = ~PC1, y = ~PC2, type = 'scatter', mode = 'markers',
                   marker = list(color = colores),
                   showlegend = FALSE, hoverinfo = "none")
  
  morado_data <- as.data.frame(pca.data[rownames(pca.data) %in% data_pool$Sample, ])
  synth_1 <- data.frame("PC1"= pca.data[rownames(pca.data) %in% synthetic[grepl("Trisomy", synthetic$Sample),]$Sample, ][,1], 
                        "PC2"= pca.data[rownames(pca.data) %in% synthetic[grepl("Trisomy", synthetic$Sample),]$Sample, ][,2], 
                        "PC3"= pca.data[rownames(pca.data) %in% synthetic[grepl("Trisomy", synthetic$Sample),]$Sample, ][,3])
  rownames(synth_1) <- synthetic_signif[grepl("Trisomy", synthetic_signif$Sample),]$Sample
  synth_2 <- data.frame("PC1"= pca.data[rownames(pca.data) %in% synthetic[grepl("Monosomy", synthetic$Sample),]$Sample, ][,1], 
                        "PC2"= pca.data[rownames(pca.data) %in% synthetic[grepl("Monosomy", synthetic$Sample),]$Sample, ][,2], 
                        "PC3"= pca.data[rownames(pca.data) %in% synthetic[grepl("Monosomy", synthetic$Sample),]$Sample, ][,3])
  rownames(synth_2) <- synthetic_signif[grepl("Monosomy", synthetic_signif$Sample),]$Sample
  
  morado_data$hover_text <- paste("Sample: ", rownames(morado_data), 
                                  "<br>PC1: ", round(morado_data$PC1, 2), 
                                  "<br>PC2: ", round(morado_data$PC2, 2))
  synth_1$hover_text <- paste("Sample: ", rownames(synth_1), 
                              "<br>PC1: ", round(synth_1$PC1, 2), 
                              "<br>PC2: ", round(synth_1$PC2, 2))
  synth_2$hover_text <- paste("Sample: ", rownames(synth_2), 
                              "<br>PC1: ", round(synth_2$PC1, 2), 
                              "<br>PC2: ", round(synth_2$PC2, 2))
  
  fig <- fig %>% add_trace(
    data = morado_data, x = ~PC1, y = ~PC2, type = "scatter", mode = "markers+text", textposition = "top right", 
    textfont = list(color = "purple", size = 10), marker = list(color = "purple"), hoverinfo = "text",
    hovertext = morado_data$hover_text)
    
  fig <- fig %>% add_trace(data = synth_1, x = ~PC1, y = ~PC2, type = "scatter", mode = "markers+text", text = "",
                           textposition = "top", textfont = list(color = "blue", size = 10), 
                           marker = list(color = "blue", size = 10), hoverinfo = "text", hovertext = synth_1$hover_text)
  
  fig <- fig %>% add_trace(data = synth_2, x = ~PC1, y = ~PC2, type = "scatter", mode = "markers+text", text = "",
                           textposition = "top", textfont = list(color = "red", size = 10),
                           marker = list(color = "red", size = 10), hoverinfo = "text", hovertext = synth_2$hover_text)
  
  fig <- fig %>% layout(title = paste0("Chromosome ", gsub(".*Chr(\\d+)", "\\1", c)), xaxis = list(title = "PC1"), yaxis = list(title = "PC2"),
                            shapes = list(list(type = "line", x0 = pca.data[sample1, "PC1"], x1 = pca.data[sample1, "PC1"], y0 = min(pca.data[,2]), y1 = max(pca.data[,2]),
                                               line = list(color = "gray", dash = "dash"), xref = "x", yref = "y"),
                                          list(type = "line", x0 = pca.data[sample2, "PC1"], x1 = pca.data[sample2, "PC1"], y0 = min(pca.data[,2]), y1 = max(pca.data[,2]), 
                                               line = list(color = "gray", dash = "dash"), xref = "x", yref = "y")), 
                            annotations = list(list(x = pca.data[sample1, "PC1"], y = max(pca.data[,2]), text = signif_percentage,
                                                    showarrow = FALSE, font = list(size = 12, color = "black"), xanchor = "left", yanchor = "top"), 
                                               list(x = pca.data[sample2, "PC1"], y = max(pca.data[,2]), text = signif_percentage, 
                                                    showarrow = FALSE, font = list(size = 12, color = "black"), xanchor = "left", yanchor = "top")))
  
  plotlis[[c]] <- fig
  
  lim_inferior <- min(pca.data[sample1, "PC1"], pca.data[sample2, "PC1"])
  lim_superior <- max(pca.data[sample1, "PC1"], pca.data[sample2, "PC1"])
  fuera <- as.data.frame(pca.data)[pca.data[,1] < lim_inferior | pca.data[,1] > lim_superior, ]
  samples <- rownames(fuera)
  for(s in samples){
    if(length(grep("Monosomy",s))==0 & length(grep("Trisomy",s))==0){
      altered_aneu[nrow(altered_aneu)+1,] <- c(s, c)
    }
  }
}

## Sex chormosomes

data_sex <- rbind(data_control, data_pool)

norms <- subset(best_norms, best_norms$Chromosome=="ChrX|Y")
top_norm <- 1

altered_sexual <- data.frame(matrix(nrow = 0, ncol = 2))
colnames(altered_sexual) <- c("Sample", "Infered_Kariotype")

for(n in norms$Normalization){
  if(n=="Norm_all"){
      Norm <- data.frame("Sample" = data_sex$Sample, 
                                    "X_norm" = data_sex$ChrX_mean/data_sex$autosomes_mean, 
                                    "Y_norm" = data_sex$ChrY_mean/data_sex$autosomes_mean)
    } else{
      Norm <- data.frame("Sample" = data_sex$Sample, 
                                    "X_norm" = data_sex$ChrX_mean/data_sex[[paste0(strsplit(n,"_")[[1]][2], "_mean")]], 
                                    "Y_norm" = data_sex$ChrY_mean/data_sex[[paste0(strsplit(n,"_")[[1]][2], "_mean")]])
    }
  
  Norm <- Norm %>% dplyr::mutate(Y_norm = ifelse(is.nan(Y_norm), as.numeric("0.0002"),Y_norm), 
                                 Karyotype = ifelse(Y_norm < 0.25, "XX","XY"))
  Norm <- Norm %>% dplyr::mutate(Pool = ifelse(Sample %in% data_pool$Sample, "Sample", "Control"))
  
  XX <- subset(Norm, Norm$Karyotype=="XX"&Norm$Pool=="Control")
  XX_Q1_X <- quantile(XX$X_norm, 0.25)
  XX_Q3_X <- quantile(XX$X_norm, 0.75)
  XX_Q1_Y <- quantile(XX$Y_norm, 0.25)
  XX_Q3_Y <- quantile(XX$Y_norm, 0.75)
  XX_IQR_value_X <- IQR(XX$X_norm)
  XX_IQR_value_Y <- IQR(XX$Y_norm)
  threshold_Xlow_XX <- (XX_Q1_X - params$threshold_signif * XX_IQR_value_X)
  threshold_Xupp_XX <- (XX_Q3_X + params$threshold_signif * XX_IQR_value_X)
  threshold_Ylow_XX <- (XX_Q1_Y - params$threshold_signif * XX_IQR_value_Y)
  threshold_Yupp_XX <- (XX_Q3_Y + params$threshold_signif * XX_IQR_value_Y)
  
  XY <- subset(Norm, Norm$Karyotype=="XY"&Norm$Pool=="Control")
  XY_Q1_X <- quantile(XY$X_norm, 0.25)
  XY_Q3_X <- quantile(XY$X_norm, 0.75)
  XY_Q1_Y <- quantile(XY$Y_norm, 0.25)
  XY_Q3_Y <- quantile(XY$Y_norm, 0.75)
  XY_IQR_value_X <- IQR(XY$X_norm)
  XY_IQR_value_Y <- IQR(XY$Y_norm)
  threshold_Xlow_XY <- (XY_Q1_X - params$threshold_signif * XY_IQR_value_X)
  threshold_Xupp_XY <- (XY_Q3_X + params$threshold_signif * XY_IQR_value_X)
  threshold_Ylow_XY <- (XY_Q1_Y - params$threshold_signif * XY_IQR_value_Y)
  threshold_Yupp_XY <- (XY_Q3_Y + params$threshold_signif * XY_IQR_value_Y)
  
  samples <- subset(Norm, Norm$Pool=="Sample")
  
  for(s in 1:nrow(samples)){
    samp <- samples[s,]
    if(samp$X_norm>threshold_Xupp_XX&samp$Y_norm<threshold_Yupp_XX){
      samp$Karyotype <- "XXX"
      altered_sexual[nrow(altered_sexual)+1,] <- samp[,c("Sample", "Karyotype")]
    }
    if(samp$X_norm>threshold_Xlow_XX&samp$Y_norm>threshold_Yupp_XX){
      samp$Karyotype <- "XXY"
      altered_sexual[nrow(altered_sexual)+1,] <- samp[,c("Sample", "Karyotype")]
    }
    if(samp$X_norm<threshold_Xlow_XX&samp$X_norm>threshold_Xupp_XY&samp$Y_norm>threshold_Yupp_XX){
      samp$Karyotype <- "M-XXY"
      altered_sexual[nrow(altered_sexual)+1,] <- samp[,c("Sample", "Karyotype")]
    }
    if(samp$X_norm<threshold_Xupp_XY&samp$Y_norm<threshold_Yupp_XX){
      samp$Karyotype <- "X0"
      altered_sexual[nrow(altered_sexual)+1,] <- samp[,c("Sample", "Karyotype")]
    }
    if(samp$X_norm<threshold_Xlow_XX&samp$X_norm>threshold_Xupp_XY&samp$Y_norm<threshold_Yupp_XX){
      samp$Karyotype <- "M-X0"
      altered_sexual[nrow(altered_sexual)+1,] <- samp[,c("Sample", "Karyotype")]
    }
    if(samp$Y_norm<threshold_Ylow_XY&samp$X_norm<threshold_Xupp_XY&samp$Y_norm>threshold_Yupp_XX){
      samp$Karyotype <- "Yp"
      altered_sexual[nrow(altered_sexual)+1,] <- samp[,c("Sample", "Karyotype")]
    }
    if(samp$X_norm<threshold_Xupp_XY&samp$Y_norm>threshold_Yupp_XY){
      samp$Karyotype <- "XYY"
      altered_sexual[nrow(altered_sexual)+1,] <- samp[,c("Sample", "Karyotype")]
    }
  }
  
  plot <- plot_ly(
        data = Norm,
        x = ~X_norm,
        y = ~Y_norm,
        type = 'scatter',
        mode = 'markers',
        symbol = ~Karyotype,
        symbols = c("circle", "square", "diamond", "cross", 
                    "x", "triangle-up", "triangle-down", "star"),
        marker = list(size = 10, color = "#a19f9f"),
        showlegend = TRUE,
        legendgroup = "shape", 
        hoverinfo = "none"
      )

      morado_data <- subset(Norm, Sample %in% data_pool$Sample)
      morado_data$hover_text <- paste(
        "Sample: ", morado_data$Sample, 
        "<br>X_norm: ", round(morado_data$X_norm, 2), 
        "<br>Y_norm: ", round(morado_data$Y_norm, 2)
      )

      plot <- plot %>% add_trace(
        data = morado_data,
        x = ~X_norm,
        y = ~Y_norm,
        type = "scatter",
        mode = "markers+text",
        marker = list(color = "purple", size = 10),
        hoverinfo = "text",
        hovertext = ~hover_text,
        showlegend = FALSE
      )

      plot <- plot %>% layout(title = paste0("Samples normalized using ", strsplit(n,"_")[[1]][2]),
                                xaxis = list(title = "Normalized X"),
                                yaxis = list(title = "Normalized Y"), 
                                legend=list(title=list(text='<b> Karyotypes </b>')))
      
      plot <- plot %>% layout(
        shapes = list(
          list(type = "line", x0 = threshold_Xlow_XX, x1 = threshold_Xlow_XX, 
               y0 = 0, y1 = 1, yref = "paper", line = list(color = "gray", dash = "dash")),
          list(type = "line", x0 = threshold_Xupp_XX, x1 = threshold_Xupp_XX, 
               y0 = 0, y1 = 1, yref = "paper", line = list(color = "gray", dash = "dash")),
          list(type = "line", x0 = threshold_Xlow_XY, x1 = threshold_Xlow_XY, 
               y0 = 0, y1 = 1, yref = "paper", line = list(color = "gray", dash = "dash")),
          list(type = "line", x0 = threshold_Xupp_XY, x1 = threshold_Xupp_XY, 
               y0 = 0, y1 = 1, yref = "paper", line = list(color = "gray", dash = "dash")),
          list(type = "line", x0 = 0, x1 = 1, xref = "paper", 
               y0 = threshold_Ylow_XX, y1 = threshold_Ylow_XX, line = list(color = "gray", dash = "dash")),
          list(type = "line", x0 = 0, x1 = 1, xref = "paper", 
               y0 = threshold_Yupp_XX, y1 = threshold_Yupp_XX, line = list(color = "gray", dash = "dash")), 
          list(type = "line", x0 = 0, x1 = 1, xref = "paper", 
               y0 = threshold_Ylow_XY, y1 = threshold_Ylow_XY, line = list(color = "gray", dash = "dash")),
          list(type = "line", x0 = 0, x1 = 1, xref = "paper", 
               y0 = threshold_Yupp_XY, y1 = threshold_Yupp_XY, line = list(color = "gray", dash = "dash"))
        )
      )
      
      plotlis[[paste0("sex_", top_norm)]] <- plot
  
  top_norm <- top_norm + 1
  
}

saveRDS(plotlis, file = paste0(params$output_dir, "Aneup_plotlis.rds"))

```

This report contains the results of the detection of autosomal and sexual anueploidies based on mean values of chromosome coverage of the different chromosomes.

## Autosomal aneuploidies {.tabset}

```{r, echo=FALSE}

if(nrow(altered_aneu)==0){
  
  write.csv(altered_aneu, file = paste0(params$output_dir, "AutosomesAneup_WARNING.csv"), row.names = F)
  
  altered_aneu[1,] <- c("    ", "    ")
  altered_aneu[2,] <- c("    ", "    ")
  
  knitr::kable(altered_aneu, row.names = FALSE)
  
} else{
  
  altered_aneu_combined <- altered_aneu %>%
  group_by(Sample) %>%
  summarise(
    Altered_autosome = paste(sort(Altered_autosome), collapse = "/"),
    .groups = "drop"
  ) %>%
  arrange(Altered_autosome)
  
  write.csv(altered_aneu_combined, file = paste0(params$output_dir, "AutosomesAneup_WARNING.csv"), row.names = F)
  
  knitr::kable(altered_aneu_combined, row.names = FALSE)
  
}

```

### Chr1

```{r, echo=FALSE}

plotlis$Chr1

```

### Chr2

```{r, echo=FALSE}

plotlis$Chr2

```

### Chr3

```{r, echo=FALSE}

plotlis$Chr3

```

### Chr4

```{r, echo=FALSE}

plotlis$Chr4

```

### Chr5

```{r, echo=FALSE}

plotlis$Chr5

```

### Chr6

```{r, echo=FALSE}

plotlis$Chr6

```

### Chr7

```{r, echo=FALSE}

plotlis$Chr7

```

### Chr8

```{r, echo=FALSE}

plotlis$Chr8

```

### Chr9

```{r, echo=FALSE}

plotlis$Chr9

```

### Chr10

```{r, echo=FALSE}

plotlis$Chr10

```

### Chr11

```{r, echo=FALSE}

plotlis$Chr11

```

### Chr12

```{r, echo=FALSE}

plotlis$Chr12

```

### Chr13

```{r, echo=FALSE}

plotlis$Chr13

```

### Chr14

```{r, echo=FALSE}

plotlis$Chr14

```

### Chr15

```{r, echo=FALSE}

plotlis$Chr15

```

### Chr16

```{r, echo=FALSE}

plotlis$Chr16

```

### Chr17

```{r, echo=FALSE}

plotlis$Chr17

```

### Chr18

```{r, echo=FALSE}

plotlis$Chr18

```

### Chr19

```{r, echo=FALSE}

plotlis$Chr19

```

### Chr20

```{r, echo=FALSE}

plotlis$Chr20

```

### Chr21

```{r, echo=FALSE}

plotlis$Chr21

```

### Chr22

```{r, echo=FALSE}

plotlis$Chr22

```

## Sexual aneuploidies {.tabset}

```{r, echo=FALSE}

if(nrow(altered_sexual)==0){
  
  write.csv(altered_sexual, file = paste0(params$output_dir, "SexualAneup_WARNING.csv"), row.names = F)
  
  altered_sexual[1,] <- c("    ", "    ")
  altered_sexual[2,] <- c("    ", "    ")
  
  knitr::kable(altered_sexual, row.names = FALSE)
  
} else{
  
  altered_sexual <- unique(altered_sexual)
  
  write.csv(altered_sexual, file = paste0(params$output_dir, "SexualAneup_WARNING.csv"), row.names = F)
  
  knitr::kable(altered_sexual, row.names = FALSE)
  
}

```

### XX-XY 1st Normalization

```{r, echo=FALSE}

plotlis$sex_1

```

### XX-XY 2nd Normalization

```{r, echo=FALSE}

plotlis$sex_2

```

### XX-XY 3rd Normalization

```{r, echo=FALSE}

plotlis$sex_3

```
